### 2024-11-29 parts/conclusion.tex
more typos

Let us end on a more general discussion regarding the status of commutative
input functions in the study of unary output polyregular functions. Throughout
the paper, we have tried to highlight that results on polyregular functions
often have the following form: a polyregular function $f \colon \Sigma^* \to
\Gamma^*$ satisfies property $X$ if and only if for all \kl{commutative}
\kl{star-free polyregular function} $g \colon \Delta^* \to \Sigma^*$, $f \circ
g$ satisfies $X$. Let us formally restate this for two earlier results in the
case of \kl{$\Rel$-polyregular functions} in
\cref{pre-compose-growth-commut:lemma,pre-compose-sf-commut:lemma}.

\begin{lemma}[restate=pre-compose-growth-commut:lemma,label=pre-compose-growth-commut:lemma]
    \proofref{pre-compose-growth-commut:lemma}
    Let $f \in \ZRat$, and $k \in \Nat$. Then,
    $f \in \ZPoly[k]$ if and only if 
    for every \kl{commutative} \kl{star-free polyregular function} $h$
            of \kl{growth rate} $l \in \Nat$,
            $(f \circ h) \in \ZPoly[k\times l]$.
\end{lemma}

@@

\begin{conjecture}
    \label{npoly-zpoly:conjecture}
    Let $k \in \Nat$.
    A function $f \in \ZPoly[k]$
    belongs to $\NPoly[k]$ if and only if
    it is \kl{$(k,\Nat)$-combinatorial}.
\end{conjecture}

>>>>

Let us end on a more general discussion regarding the status of commutative
input functions in the study of unary output polyregular functions. A
\emph{quantitative pumping argument} for polyregular function $f \colon
\Sigma^* \to \Rel$ states that $f$ has property $X$ if and only if for all
\kl{pumping pattern} $q \colon \Nat^k \to \Sigma^*$, $f \circ q$ has property
$X$. Let us formalize such a statement for \kl{growth rate} and
\kl(polyregular){aperiodicity} respectively in \cref{pre-compose-growth-commut:lemma,pre-compose-sf-commut:lemma}. Note that we generalized \kl{pumping
patterns} to \kl{commutative} \kl{star-free polyregular functions} to simplify
the statements.

\begin{lemma}[restate=pre-compose-growth-commut:lemma,label=pre-compose-growth-commut:lemma]
    \proofref{pre-compose-growth-commut:lemma}
    Let $f \in \ZRat$, and $d \in \Nat$. Then,
    $f \in \ZPoly[d]$ if and only if 
    for every \kl{commutative} \kl{star-free polyregular function} $h$
            of \kl{growth rate} $l \in \Nat$,
            $(f \circ h) \in \ZPoly[d \times l]$.
\end{lemma}

@@

\begin{conjecture}
    \label{npoly-zpoly:conjecture}
    Let $f \in \ZPoly$. Then, $f \in \NPoly$ if and only if for every
    \kl{commutative} \kl{star-free polyregular function} $h$,
    $(f \circ h) \in \NPoly$.
\end{conjecture}

### 2024-11-29 parts/starfree.tex
more typos

In this section, we leverage the decidability results of \cref{polynomials:sec}
to decide membership in $\NPoly$
inside $\ZPoly$ and membership in $\NSF$ inside $\NPoly$, both under the extra
assumption of \kl{commutativity}. 
To characterize $\NPoly$ inside $\ZPoly$ we introduce the notion of
\kl{$(k,\Nat)$-combinatorial} function
(\cref{k-combinatorial:def}), following the spirit of previous
characterizations of subclasses of $\ZPoly$ in terms of \emph{polynomial
pumping arguments} \cite{DOUE21,DOUE22,CDTL23}.

@@

\begin{theorem}
    \label{decidable-n-poly:thm}
    Let $k \in \Nat$, and $f \in \ZPoly[k]$ be \kl{commutative}.
    Then, the following are equivalent:
    \begin{enumerate}
        \item \label{f-combinatorial:item} $f$ is \kl{$(k,\Nat)$-combinatorial},
        \item \label{f-npoly-combi:item} $f \in \NPoly[k]$,
    \end{enumerate}
    Furthermore, the properties are decidable,
    and conversions effective.
\end{theorem}
\begin{proof}
    The implication
    \cref{f-npoly-combi:item}
    $\implies$ \cref{f-combinatorial:item} immediately follows
    from 
    \cref{n-poly-combinatorics:lem} and \cref{decide-rat-poly-npoly:cor}.
    For the converse implication, we use \cref{decompose-polynomial:lem} 
    effectively compute an $\omega \in \Nat$ and \kl{integer binomial polynomials}
    $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^k}$.

@@

    To conclude, one can compute $f$ by first
    computing the \kl{$\omega$-type} of the input which is possible using a \kl{polyregular function},
    and then computing the suitable \kl{strongly natural binomial polynomials}
    which is possible in $\NPoly$
    thanks to \cref{decide-rat-poly-npoly:cor}. The resulting composition 
    belongs to $\NPoly$ thanks to
    \cref{stability-polyregular:lemma}.

    Let us now argue that the properties are decidable. Indeed, consider a
    function $f \in \ZPoly[k]$. Again, using
    \cref{decompose-polynomial:lem}, we can effectively compute the
    $\omega \in \Nat$ and the \kl{integer binomial polynomials}
    $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^k}$. If all of those polynomials
    are \kl{strongly natural binomial polynomials} (which is 
    decidable by \cref{decide-rat-poly-npoly:cor}),
    then $f \in \NPoly[k]$ as
    argued above. Conversely, if $f \in \NPoly[k]$, then $f$ is
    \kl{$(k,\Nat)$-combinatorial} by \cref{n-poly-combinatorics:lem}, and we
    argued that this implies that $P_{(S, \vec{r})}$ is a \kl{strongly natural
    binomial polynomial} for all $(S, \vec{r}) \in \ModuloTypes[\omega]^k$.
\end{proof}

@@

\begin{definition}[Ultimately polynomial]
    \label{ultimately-polynomial:def}
    Let $\Sigma$ be a finite alphabet. 
    A function $f \colon \Sigma^* \to \Rel$
    is \intro{ultimately polynomial}
    when there exists $N_0 \in \Nat$ such that
    for all $n \in \Nat$,
    for all \kl{pumping pattern} $q \colon \Nat^n \to \Sigma^*$,
    there exists a polynomial $P \in \Rat[X_1, \dots, X_n]$
    such that:
    \begin{equation*}
        f \circ p = P
        \quad 
        \text{ over } (\Nat_{\geq N_0})^n
        \quad .
    \end{equation*}
\end{definition}

@@

\begin{theorem}
    \label{zsf-npoly-nsf:thm}
    Let $\Sigma$ be a finite alphabet, 
    and $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Nat$-polyregular function}.
    Then, the following are equivalent:
    \begin{enumerate}
        \item \label{zsf-npoly-nsf:thm:up:item} $f$ is \kl{ultimately polynomial},
        \item \label{zsf-npoly-nsf:thm:zsf:item} $f \in \ZSF$,
        \item \label{zsf-npoly-nsf:thm:nsf:item} $f \in \NSF$.
    \end{enumerate}
    Furthermore, membership is decidable and conversions are effective.
    \proofref{zsf-npoly-nsf:thm}
\end{theorem}
\begin{proof}
    The implication \cref{zsf-npoly-nsf:thm:nsf:item} $\Rightarrow$
    \cref{zsf-npoly-nsf:thm:zsf:item} is immediate since $\NSF \subseteq \ZSF$.
    Furthermore,
    \cref{zsf-npoly-nsf:thm:zsf:item} implies \cref{zsf-npoly-nsf:thm:up:item}
    following previous results for \kl{star-free $\Rel$-polyregular functions}
    \cite{CDTL23}.

    For the implication \cref{zsf-npoly-nsf:thm:up:item} $\Rightarrow$
    \cref{zsf-npoly-nsf:thm:nsf:item}, let us assume that $f$ is \kl{ultimately
    polynomial}. We prove the result by induction on the size of the alphabet
    $\Sigma$. By definition, there exists $N_0 \in \Nat$, and $P \in
    \Rat[\seqof{X_a}{a \in \Sigma}]$ such that: \begin{equation*} f\left(
        \prod_{a \in \Sigma} a^{X_a} \right) = P(\seqof{X_a}{a \in \Sigma})
        \quad \text{ when } \forall a \in \Sigma, X_a \geq N_0 \quad .
    \end{equation*} It is clear that $P$ is \kl{represented} by a
    \kl{$\Nat$-polyregular function}, namely, $f$ partially applied to the word
    $\prod_{a \in \Sigma} a^{N_0}$, and therefore is in fact represented by a
    \kl{star-free $\Nat$-polyregular function} thanks to
    \cref{decide-rat-poly-npoly:cor}.
    For every letter $a \in \Sigma$ and number $0 \leq n \leq N_0$, there exists, by induction hypothesis,
    a \kl{star-free $\Nat$-polyregular function} $g_a^n$ that represents the
    function $f_{a^n} \colon (\Sigma \setminus \set{a})^* \to \Rel$
    that maps $u \in (\Sigma \setminus \set{a})^*$ to $f(a^n u)$.

    Let us now define the function $g \colon \Sigma^* \to \Rel$ as follows:
    \begin{equation*}
        g \colon w \mapsto \begin{cases}
            g_{a^n} (w) & \text{ if } \card[a]{w} = n \text{ for some } a \in \Sigma \text{ and } n \leq N_0 \\
            P(\seqof{\card[a]{w}}{a \in \Sigma}) & \text{ otherwise }
        \end{cases}
    \end{equation*}
    We conclude that $g$ is a \kl{star-free $\Nat$-polyregular function}, and by definition
    $f = g$.
\end{proof}

>>>>

In this section, we leverage the decidability results of \cref{polynomials:sec}
to decide membership in $\NPoly$
inside $\ZPoly$ and membership in $\NSF$ inside $\NPoly$, both under the extra
assumption of \kl{commutativity}. 
To characterize $\NPoly$ inside $\ZPoly$ we introduce the notion of
\kl{$(k,\Nat)$-combinatorial} function
(\cref{k-combinatorial:def}), following the spirit of revious
characterizations of subclasses of $\ZPoly$ in terms of \emph{polynomial
pumping arguments} \cite{DOUE21,DOUE22,CDTL23}.

@@

\begin{theorem}
    \label{decidable-n-poly:thm}
    Let $k,d \in \Nat$, and $f \in \ZPoly[d]$ be \kl{commutative}
    over an alphabet of size $k$.
    Then, the following are equivalent:
    \begin{enumerate}
        \item \label{f-combinatorial:item} $f$ is \kl{$(k,\Nat)$-combinatorial},
        \item \label{f-npoly-combi:item} $f \in \NPoly[d]$,
    \end{enumerate}
    Furthermore, the properties are decidable,
    and conversions effective.
\end{theorem}
\begin{proof}
    The implication
    \cref{f-npoly-combi:item}
    $\implies$ \cref{f-combinatorial:item} immediately follows
    from 
    \cref{n-poly-combinatorics:lem} and \cref{decide-rat-poly-npoly:cor}.
    For the converse implication, we use \cref{decompose-polynomial:lem} 
    effectively compute an $\omega \in \Nat$ and \kl{integer binomial polynomials}
    $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^k}$.

@@

    To conclude that $f \in \NPoly$, one can compute $f$ by first computing
    the \kl{$\omega$-type} of the input which is possible using a
    \kl{polyregular function}, and then computing
    the suitable \kl{strongly natural binomial polynomials} which is possible
    in $\NPoly$ thanks to \cref{decide-rat-poly-npoly:cor}. The
    resulting composition belongs to $\NPoly$ thanks to
    \cref{stability-polyregular:lemma},
    and we conclude that $f \in \NPoly[d]$ because
    it has \kl{growth rate} at most $d$
    (\cref{polyregular-polynomial-growth:lemma}).

    Let us remark that we actually proved that the function $f \in \ZPoly[d]$
    belongs to $\NPoly[d]$
    if and only if its corresponding polynomials
    $P_{(S, \vec{r})}$ are \kl{strongly natural binomial polynomials} for all
    $(S, \vec{r}) \in \ModuloTypes[\omega]^k$. Because 
    \cref{decompose-polynomial:lem} is effective, this $\omega \in \Nat$ and polynomials
    $P_{(S, \vec{r})}$ can be computed from $f$, 
    and one can decide 
    whether a polynomial is \kl{strongly natural binomial polynomial} using
    \cref{decide-rat-poly-npoly:cor}. Hence, one can decide whether
    $f \in \NPoly[d]$.
\end{proof}

@@

\begin{definition}[Ultimately polynomial]
    \label{ultimately-polynomial:def}
    Let $\Sigma$ be a finite alphabet. 
    A function $f \colon \Sigma^* \to \Rel$
    is \intro{ultimately polynomial}
    when there exists $N_0 \in \Nat$ such that
    for all $k \in \Nat$,
    for all \kl{pumping pattern} $q \colon \Nat^k \to \Sigma^*$,
    there exists a polynomial $P \in \Rat[X_1, \dots, X_k]$
    such that:
    \begin{equation*}
        f \circ q = P
        \quad 
        \text{ over } (\Nat_{\geq N_0})^k
        \quad .
    \end{equation*}
\end{definition}

@@

\begin{theorem}
    \label{zsf-npoly-nsf:thm}
    Let $\Sigma$ be a finite alphabet, 
    and $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Nat$-polyregular function}.
    Then, the following are equivalent:
    \begin{enumerate}
        \item \label{zsf-npoly-nsf:thm:up:item} $f$ is \kl{ultimately polynomial},
        \item \label{zsf-npoly-nsf:thm:zsf:item} $f \in \ZSF$,
        \item \label{zsf-npoly-nsf:thm:nsf:item} $f \in \NSF$.
    \end{enumerate}
    Furthermore, membership is decidable and conversions are effective.
    \proofref{zsf-npoly-nsf:thm}
\end{theorem}
\begin{proof}
    The implication \cref{zsf-npoly-nsf:thm:nsf:item} $\Rightarrow$
    \cref{zsf-npoly-nsf:thm:zsf:item} is immediate since $\NSF \subseteq \ZSF$.
    Furthermore,
    \cref{zsf-npoly-nsf:thm:zsf:item} implies \cref{zsf-npoly-nsf:thm:up:item}
    following previous results for \kl{star-free $\Rel$-polyregular functions}
    \cite[Theorem V.13]{CDTL23}.

    For the implication \cref{zsf-npoly-nsf:thm:up:item} $\Rightarrow$
    \cref{zsf-npoly-nsf:thm:nsf:item}, let us assume that $f$ is \kl{ultimately
    polynomial}. We prove the result by induction on the size of the alphabet
    $\Sigma$. By definition, there exists $N_0 \in \Nat$, and $P \in
    \Rat[\seqof{X_a}{a \in \Sigma}]$ such that:

    \begin{equation*} f\left(
        \prod_{a \in \Sigma} a^{x_a} \right)
        = 
        P(\seqof{x_a}{a \in \Sigma})
        \quad \forall \vec{x} \in (\Nat_{\geq N_0})^{\Sigma} \quad .
    \end{equation*}

    It is clear that $\translate{N_0}(P)$ is \kl{represented} by an
    \kl{$\Nat$-polyregular function}, namely, $f_u \colon w \mapsto f(uw)$
    where $u \defined \prod_{a \in \Sigma} a^{N_0}$, and is therefore represented by a
    \kl{star-free $\Nat$-polyregular function} thanks to
    \cref{decide-rat-poly-npoly:cor}.
    For every letter $a \in \Sigma$ and number $0 \leq n \leq N_0$, there exists, by induction hypothesis,
    a \kl{star-free $\Nat$-polyregular function} $g_{a^n}$ that \kl{represents} the
    function $f_{a^n} \colon (\Sigma \setminus \set{a})^* \to \Rel$
    that maps $w \in (\Sigma \setminus \set{a})^*$ to $f(a^n w)$.

    Let us now define the function $g \colon \Sigma^* \to \Rel$ as follows:
    \begin{equation*}
        g \colon w \mapsto \begin{cases}
            g_{a^n} (w) & \text{ if } \card[a]{w} = n \text{ for some } a \in \Sigma \text{ and } n \leq N_0 \\
            \translate{N_0}(P)(\seqof{\card[a]{w} - N_0}{a \in \Sigma}) & \text{ otherwise }
        \end{cases}
    \end{equation*}
    We conclude that $g$ is a \kl{star-free $\Nat$-polyregular function}, and by definition
    $f = g$.
\end{proof}
